---
title: "`tinytable`"
---

```{r}
# library(tinytable)
pkgload::load_all()
options(tt_placement = "H")
x <- mtcars[1:4, 1:4]
```

```{r}
tt(x) |> style_cell(2, 2, bg = "pink !important")
```


# Width

Extendable columns:

```{r}
tt(x,
  bootstrap_class = "table table-dark",
  tabularray_extendable = TRUE,
  tabularray_inner = "width=.7\\linewidth")
```

# Transform all entries

By default, all functions transform all cells. For example, calling `style_column()` without specifying column numbers will apply the transformation to all columns:

TODO: HTML Broken

```{r}
tt(x) |>
  style_column(bg = "pink")
```

```{r}
tt(x) |>
  style_row(bg = "pink")
```

# Columns

```{r}
tt(x) |>
  style_column(j = 2:3, fg = "white", bg = "olive", cmd="\\texttt") |>
  style_row(i = 2:3, fg = "white", bg = "olive", cmd="\\texttt")
```

# Rows

```{r}
tt(x) |>
  style_row(i = 2:3, fg = "blue6", bg = "pink", font="\\bfseries")
```

# Cells

## Style

```{r}
tt(x) |>
  style_column(j = 1, halign = "c") |>
  style_cell(i = 2:3, j = 1, fg = "red") |>
  style_cell(i = 4, j = 1, bg = "red")
```

## Spanning cells

```{r}
tt(x) |>
  style_cell(i = 2, j = 1, r = 3, c = 2, bg = "pink")
```

## Special characters

TODO: this does not work

```{r, eval=FALSE}
k <- data.frame(a = c("a_b", "d$c"))
tt(k, tabularray_inner = "cmd=\\fakeverb")
```


# Caption and labels

# Math

```{r}
tt(x, output = "html") |>
  style_column(j = 2:3, fg = "white", bg = "olive", cmd="\\texttt") |>
  style_row(i = 2:3, fg = "white", bg = "pink", cmd="\\texttt")
```

# HTML-specific features

## Bootstrap classes

<https://getbootstrap.com/docs/5.3/content/tables/>

```{r, eval = knitr::is_html_output()}
tt(x, bootstrap_class = "table table-dark")
```

```{r, eval = knitr::is_html_output()}
tt(x, bootstrap_class = "table table-striped")
```

## Custom CSS

```{css}
.table-custom th, .table-custom td {
  text-align: right;
  background-color: pink !important;
}
```

```{r, eval = knitr::is_html_output()}
tt(x, bootstrap_class = "table table-custom")
```


