# FAQ

```{=html}
<script id="MathJax-script" async src="https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"></script>
<script>
MathJax = {
  tex: {
    inlineMath: [['$', '$'], ['\\(', '\\)']]
  },
  svg: {
    fontCache: 'global'
  }
};
</script>
```

## LaTeX

### Preamble

`tinytable` uses the `tabularray` package from your LaTeX distribution to draw tables. `tabularray`, in turn, provides special `tblr`, `talltblr`, and `longtblr` environments to display tabular data.

When rendering a document from Quarto or Rmarkdown directly to PDF, `tinytable` will populate the LaTeX preamble automatically with all the required packages (except when code chunks are cached). For standalone LaTeX documents, these commands should be inserted in the preamble manually:

```latex
\usepackage{tabularray}
\usepackage{float}
\usepackage{graphicx}
\usepackage[normalem]{ulem}
\UseTblrLibrary{booktabs}
\UseTblrLibrary{siunitx}
\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
```

### `setspace`

Some users have encountered unexpected spacing behavior when generating tables that are *not* wrapped in a `\begin{table}` environment (ex: `multipage` or raw `tblr`).

One issue stems from the fact that the `\begin{table}` environment resets any spacing commands in the preamble or body by default, such as:

```tex
\usepackage{setspace}
\doublespacing
```

This means that when using `theme_tt("multipage")` ---which does not wrap the table in a `table` environment--- the spacing is *not* reset, and tables are double spaced. This is not a bug, since double-spacing is in fact what the user requested. Nevertheless, the behavior can seem surprising for those used to the automagical `table` environment spacing reset.

One workaround is to add the following to the document preamble when using multipage/longtblr:

```tex
\usepackage{etoolbox}
\AtBeginEnvironment{longtblr}{\begin{singlespacing}}
\AtEndEnvironment{longtblr}{\end{singlespacing}}
```

Example Quarto doc:

  
````md
---
title: longtblr and setspacing
format:
  pdf:
    include-in-header:
      - text: |
         % Tinytable preamble
         \usepackage{tabularray}
         \usepackage{float}
         \usepackage{graphicx}
         \usepackage{codehigh}
         \usepackage[normalem]{ulem}
         \UseTblrLibrary{booktabs}
         \UseTblrLibrary{siunitx}
         \newcommand{\tinytableTabularrayUnderline}[1]{\underline
         {#1}}
         \newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
         \NewTableCommand{\tinytableDefineColor}[3]{\definecolor{
         #1}{#2}{#3}}
         % Spacing Commands
         \usepackage{setspace}
         \doublespacing
         % Fix Spacing in longtblr
         \usepackage{etoolbox}
         \AtBeginEnvironment{longtblr}{\begin{singlespacing}}
         \AtEndEnvironment{longtblr}{\end{singlespacing}}
---

```{=latex}
\begin{longtblr}[         %% tabularray outer open
]                     %% tabularray outer close
{                     %% tabularray inner open
colspec={Q[]Q[]Q[]Q[]},
}                     %% tabularray inner close
\toprule
foo & bar & baz \\
foo & bar & baz \\
foo & bar & baz \\
\bottomrule
\end{longtblr}
```
````

### Global styles

`tabularray` allows very powerful styling and themeing options. See [the reference manual](https://ctan.org/pkg/tabularray) for more information.

For example, you can change the size of footnotes in all tables of a document with:

````{verbatim}
---
format: 
    pdf:
        keep-tex: true
        header-includes: |
            \SetTblrStyle{foot}{font=\LARGE}
---

```{r}
library(tinytable)
library(magrittr)
tt(head(iris), notes = "Blah blah")
```
````


### Beamer

Due to [a bug in the upstream package `rmarkdown`](https://github.com/rstudio/rmarkdown/issues/2478), Quarto or Rmarkdown presentations compiled to Beamer cannot include adequate package loading commands in the preamble automatically. This bug prevents `tinytable::usepackage_latex()` from modifying the preamble. Here's a workaround.

Save this LaTeX code as `preamble.tex`:

```latex
\RequirePackage{tabularray}
\RequirePackage{booktabs}
\RequirePackage{float}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\UseTblrLibrary{booktabs}
\UseTblrLibrary{siunitx}
\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
```

Then, load `preamble.tex` in your YAML header:

````yaml
---
output: 
  beamer_presentation:
    includes:
      in_header: preamble.tex
---
````

With these changes, the table should appear with colors as expected.


## Typst

### Captions and labels

Due to a quirk in Quarto [reported here,](https://github.com/quarto-dev/quarto-cli/issues/9369) captions in Typst and Quarto documents need to be specified using *both* the `tbl-cap` and the `label` chunk option:

````{verbatim}
```{r}
#| tbl-cap: "blah blah blah"
#| label: tbl-blah
tinytable::tt(head(iris, 5))
```
````

### Multi-page long tables

The Typst tables created by `tinytable` are automatically broken across pages with repeated headers. However, in Quarto documents, the Quarto software wraps tables in an non-breakable `#figure` environment. This can break the display of long tables. One solution is to use a raw Typst code block to set Figures to be breakable:

````{verbatim}
---
format: typst
---

```{=typst}
#show figure: set block(breakable: true)
```

```{r}
#| tbl-cap: "blah blah blah"
#| label: tbl-blah
library(tinytable)
tt(head(iris, 50))
```
````


## Markdown

### `style_tt()` does not apply to row headers

This is an important limitation, but it is difficult to get around. See this issue for discussion: https://github.com/vincentarelbundock/tinytable/issues/125

Users can use markdown styling directly in `group_tt()` to circumvent this. This is documented in the tutorial.

### `rowspan` and `colspan`

These arguments are already implemented in the form of "pseudo-spans", meaning that we flush the content of adjacent cells, but do not modify the row or column borders. This is probably adequate for most needs. 

One alternative would be to remove line segments in finalize_grid(). I tried this but it is tricky and the results were brittle, so I rolled it back. I'm open to considering a PR if someone wants to contribute code, but please discuss the feature design in an issue with me before working on this.

## Quarto

### Custom crossref styles

In `Quarto`, it is possible to create a custom crossref type for things like appendix tables. One challenge, is that LaTeX will not allow users to nest a `tblr` environment, inside a `table` environment, inside the new environment that `Quarto` creates for the crossref. Therefore, when rendering a table to LaTeX/PDF, it is important to drop the `\begin{table}` environment. This can be done using the `theme_tt()` function.

In the example below, we call `theme_tt()` explicitly for a single table, but the [themes vignette](https://vincentarelbundock.github.io/tinytable/vignettes/theme.html) shows how to set a global theme using the `tinytable_tt_theme` option.

````{verbatim}
---
title: "Crossref Example"
format:
  pdf: default
  html: default
crossref:
  custom:
    - kind: float
      key: apptbl
      latex-env: apptbl
      reference-prefix: Table A
      space-before-numbering: false
      latex-list-of-description: Appendix Table
apptbl-cap-location: top
---

See @apptbl-testing

::: {#apptbl-testing}

```{r}
library(tinytable)

tt(mtcars[1:5,]) |> theme_tt("tabular", style = "tabularray")
```

Caption goes here.

:::
````
