# FAQ

## LaTeX

### `setspace`

Some users have encountered unexpected spacing behavior when generating tables that are *not* wrapped in a `\begin{table}` environment (ex: `multipage` or raw `tblr`).

One issue stems from the fact that the `\begin{table}` environment resets any spacing commands in the preamble or body by default, such as:

```tex
\usepackage{setspace}
\doublespacing
```

This means that when using `theme_tt("multipage")` ---which does not wrap the table in a `table` environment--- the spacing is *not* reset, and tables are double spaced. This is not a bug, since double-spacing is in fact what the user requested. Nevertheless, the behavior can seem surprising for those used to the automagical `table` environment spacing reset.

One workaround is to add the following to the document preamble when using multipage/longtblr:

```tex
\usepackage{etoolbox}
\AtBeginEnvironment{longtblr}{\begin{singlespacing}}
\AtEndEnvironment{longtblr}{\end{singlespacing}}
```

Example Quarto doc:

  
````md
---
title: longtblr and setspacing
format:
  pdf:
    include-in-header:
      - text: |
         % Tinytable preamble
         \usepackage{tabularray}
         \usepackage{float}
         \usepackage{graphicx}
         \usepackage{codehigh}
         \usepackage[normalem]{ulem}
         \UseTblrLibrary{booktabs}
         \UseTblrLibrary{siunitx}
         \newcommand{\tinytableTabularrayUnderline}[1]{\underline
         {#1}}
         \newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
         \NewTableCommand{\tinytableDefineColor}[3]{\definecolor{
         #1}{#2}{#3}}
         % Spacing Commands
         \usepackage{setspace}
         \doublespacing
         % Fix Spacing in longtblr
         \usepackage{etoolbox}
         \AtBeginEnvironment{longtblr}{\begin{singlespacing}}
         \AtEndEnvironment{longtblr}{\end{singlespacing}}
---

```{=latex}
\begin{longtblr}[         %% tabularray outer open
]                     %% tabularray outer close
{                     %% tabularray inner open
colspec={Q[]Q[]Q[]Q[]},
}                     %% tabularray inner close
\toprule
foo & bar & baz \\
foo & bar & baz \\
foo & bar & baz \\
\bottomrule
\end{longtblr}
```
````


### Beamer

Due to [a bug in the upstream package `rmarkdown`](https://github.com/rstudio/rmarkdown/issues/2478), Quarto or Rmarkdown presentations compiled to Beamer cannot include adequate package loading commands in the preamble automatically. This bug prevents `tinytable::usepackage_latex()` from modifying the preamble. Here's a workaround.

Save this LaTeX code as `preamble.tex`:

```latex
\RequirePackage{tabularray}
\RequirePackage{booktabs}
\RequirePackage{float}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\UseTblrLibrary{booktabs}
\UseTblrLibrary{siunitx}
\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
```

Then, load `preamble.tex` in your YAML header:

````yaml
---
output: 
  beamer_presentation:
    includes:
      in_header: preamble.tex
---
````

With these changes, the table should appear with colors as expected.


## Typst

### Captions and labels

Due to a quirk in Quarto [reported here,](https://github.com/quarto-dev/quarto-cli/issues/9369) captions in Typst and Quarto documents need to be specified using *both* the `tbl-cap` and the `label` chunk option:

````
```{r}
#| tbl-cap: "blah blah blah"
#| label: tbl-blah
tinytable::tt(head(iris, 5))
```
````

### Multi-page long tables

The Typst tables created by `tinytable` are automatically broken across pages with repeated headers. However, in Quarto documents, the Quarto software wraps tables in an non-breakable `#figure` environment. This can break the display of long tables. One solution is to use a raw Typst code block to set Figures to be breakable:

````{verbatim}
---
format: typst
---

```{=typst}
#show figure: set block(breakable: true)
```

```{r}
#| tbl-cap: "blah blah blah"
#| label: tbl-blah
library(tinytable)
tt(head(iris, 50))
```
````