```{r}
#| include: false
library(tinytable)
options(tinytable_html_engine = "tabulator")
```

# Interactive tables

::: {.callout-warning}
## Experimental Feature
The Tabulator.js integration is experimental and the API may change in future versions. Please report any issues on [GitHub](https://github.com/vincentarelbundock/tinytable/issues).
:::

The `tinytable` package supports creating interactive tables using [the Tabulator.js library.](https://tabulator.info/) Tabulator is a powerful JavaScript library that provides features like 

* Sorting 
* Filtering
* Pagination
* Themes
* Data export
* Real-time data editing in the browser
* Accessibility features (ARIA compliant)

## Drawing, printing, and saving

To create an interactive table, use `output = "tabulator"` when printing your table:

```{r}
#| eval: false
library(tinytable)
dat <- head(iris)
tt(dat) |> print("tabulator")
```

To save the table to file, we can use the `save_tt()` function. One issue to consider, however, is that `tinytable` supports two types of HTML output: `tabulator` and `bootstrap`. To select the default HTML engine, users can set a global option:

```{r}
#| eval: false
options(tinytable_html_engine = "tabulator")
tt(dat) |> save_tt("/path/to/your/file.html")
```

In notebooks like Quarto or R markdown, `tinytable` will automatically create an HTML when appropriate. Whenn the `tinytable_html_engine` option is set to `"tabulator"`, the table will be rendered using Tabulator.js.

```{r}
dat <- data.frame(
  city = c("Montréal", "Toronto", "Vancouver"),
  salary = c(14002.22, 201399.11, 80188.38),
  random = c(1.43402, 201.399, 0.134588),
  date = as.Date(sample(1:1000, 3), origin = "1970-01-01"),
  best = c(TRUE, FALSE, FALSE)
)

tt(dat)
```

## Pagination and filtering

`tinytable` includes a built-in theme to add pagination, sorting, and filtering capabilities to a Tabulator table. This is particularly useful for large datasets. 

To apply a theme, call the `theme_tt()` function with the `"tabulator"` argument. See `?theme_tt` for a list of arguments that can be used to customize the number of pagination rows, behaviour of the search bar, and various other elements.

Try typing the letters "vir" in the search box to filter the `iris` dataset and find the Virginica flowers.

```{r}
tt(iris) |> theme_tt("tabulator", pagination = c(5, 10, 50))
```


## Format

Formatting numeric and date columns in Tabulator tables requires us to use the Javascript functionality rather than `tinytable`'s internals, if we want to preserve functionality like sorting. 

In particular, for numeric values, `format_tt()` is always set to `num_fmt="decimal"`. 

```{r}
tt(dat) |>
  format_tt(j = "salary", digits = 2, num_mark_big = ",") |>
  format_tt(j = "random", digits = 4)
```

For dates, 

```{r}
tt(dat) |> format_tt(j = "date", date = "M/d/yyyy")
```

For dates, `tabulator` uses Luxon date format tokens, not R's `strptime` format. Common patterns include:

```{r}
#| echo: false
luxon_formats <- data.frame(
  "Luxon Format" = c("M/d/yyyy", "MM/dd/yyyy", "d/M/yyyy", "dd/MM/yyyy", "yyyy-MM-dd", "MMM d, yyyy"),
  "Output Example" = c("1/15/2020", "01/15/2020", "15/1/2020", "15/01/2020", "2020-01-15", "Jan 15, 2020"),
  "Description" = c("US short format", "US padded format", "European short format", "European padded format", "ISO format", "Long month name"),
  check.names = FALSE
)

tt(luxon_formats)
```

## Style

Support for `style_tt()` is very limited in interactive tables. For now, only the `align` and `alignv` arguments of that function are supported.

```{r}
tt(dat) |> style_tt(align = "r")
```

## CSS

For more advanced styling, you can add custom CSS rules using the `css_rule` argument in `theme_tt()`. The CSS rule must include at least one `$TINYTABLE_ID` placeholder, which gets replaced with the unique table identifier to ensure styles only apply to that specific table.

```{r}
css_rule <- "
$TINYTABLE_ID .tabulator-header .tabulator-col {
  background-color: black;
  color: white;
}
"
tt(dat) |> 
  theme_tt("tabulator", css_rule = css_rule)
```

## Style sheets

Tabulator ships with multiple complete CSS style sheets. The default in `tinytable` is Bootstrap 5, but you can customize the appearance using the `theme_tt()` function (when available). Alternatives include `"default"`, `"simple"`, `"midnight"`, `"modern"`, `"site"`, `"site_dark"`, `"bootstrap3"`, `"bootstrap4"`, `"bootstrap5"`, `"semanticui"`, `"bulma"`, and `"materialize"`.

```{r}
#| eval: false
tt(iris) |> theme_tt("tabulator", stylesheet = "semanticui")
```

::: {.callout-note}
To use a theme, the HTML file must load a style sheet *globally* in the document. Unfortunately, this means that `tinytable` cannot apply a `Tabulator` style sheet to a single table in documents with multiple tables. The style that applies is always the last one loaded in the document.
:::


## Options and columns

The `theme_tt()` function accepts `options` and `columns` arguments for advanced customization of Tabulator tables.

The `options` argument allows you to override any default Tabulator configuration. The `columns` argument lets you completely customize column definitions, including formatters, styling, and behavior.

In this example, we redefine how columns are formatted.

```{r}
dat <- data.frame(
  city = c("Toronto", "Montreal", "Vancouver"),
  salary = c(75000, 68000, 82000),
  best = c(FALSE, TRUE, FALSE)
)

custom_columns <- '
[
  {
    "title": "City", 
    "field": "city"
  }, 
  {
    "title": "Best city", 
    "field": "best", 
    "formatter": "tickCross"
  }, 
  {
    "title": "Salary", 
    "field": "salary", 
    "formatter": "money", 
    "formatterParams": {"precision": 0, "symbol": "$"}
  },
]'

tt(dat) |> theme_tt("tabulator", columns = custom_columns)
```

And now we change more options, such as the layout and height of the table:

```{r}
opts <- "
  layout: 'fitColumns',
  height: '200px'
"

tt(dat) |> theme_tt("tabulator", options = opts)
```





```{r}
#| include: false
# reset the global option to Bootstrap to avoid affecting other documents
options(tinytable_html_engine = NULL)
```
