# Notebooks (Quarto, Rmarkdown, Bookdown, etc.)


## Quarto

### Custom crossref styles

In `Quarto`, it is possible to create a custom crossref type for things like appendix tables. One challenge, is that LaTeX will not allow users to nest a `tblr` environment, inside a `table` environment, inside the new environment that `Quarto` creates for the crossref. Therefore, when rendering a table to LaTeX/PDF, it is important to drop the `\begin{table}` environment. This can be done using the `theme_tt()` function.

In the example below, we call `theme_tt()` explicitly for a single table, but the [themes vignette](https://vincentarelbundock.github.io/tinytable/vignettes/theme.html) shows how to set a global theme using the `tinytable_tt_theme` option.

````{verbatim}
---
title: "Crossref Example"
format:
  pdf: default
  html: default
crossref:
  custom:
    - kind: float
      key: apptbl
      latex-env: apptbl
      reference-prefix: Table A
      space-before-numbering: false
      latex-list-of-description: Appendix Table
apptbl-cap-location: top
---

See @apptbl-testing

::: {#apptbl-testing}

```{r}
library(tinytable)

tt(mtcars[1:5,]) |> theme_tt("tabular", style = "tabularray")
```

Caption goes here.

:::
````

### Sub-captions

With version 0.4.0 of `tinytable` and versions above 1.6 of Quarto---only available [as a pre-release build at the time of writing](https://quarto.org/docs/download/prerelease.html)--- users can specify sub-captions for tables. This is done by adding a `tbl-subcap` chunk option to the table chunk. We also need to use a `tinytable` theme to remove built-in table structure and let Quarto handle the table construction. Here is a simple example document:

````{verbatim}
---
format: pdf
---

See @tbl-example, @tbl-example-1, or @tbl-example-2.

```{r}
#| label: tbl-example
#| tbl-cap: "Example"
#| tbl-subcap: 
#|   - "Cars"
#|   - "Pressure"
#| layout-ncol: 2

library(tinytable)
tt(head(cars)) |> theme_tt("tabular", style = "tabularray")
tt(head(pressure)) |> theme_tt("tabular", style = "tabularray")
```
````

### Same table, different styles

In some cases, the user wants to print a single table multiple times with different styles in a single HTML document. This will sometimes cause issues, because the `style_tt()` function inserts several javascript functions to modify the same table, thus creating conflicts. A simple solution is to change the unique ID of the table object manually. 

Consider this RevalJS slideshow in which we sequentially highlight different rows of the data frame:

````{verbatim}
---
format: revealjs
---

## Page 1

```{r}
library(tinytable)

tab <- tt(head(iris))
tab
```

## Page 2

```{r}
tab@id <- "table_01"
tab |> style_tt(i = 3, background = "skyblue")
```

## Page 3

```{r}
tab@id <- "table_02"
tab |> style_tt(i = 5, background = "orange")
```
````


## Bookdown

### Cross-references

The `bookdown` package uses a special syntax to handle cross-references, and it does not recognize `tinytable` objects as tables automatically. To include cross-references to tables, it is thus necessary to use the `caption` argument of `tinytable::tt()`, and to insert a `bookdown` label in that caption. Here is an example:


````
Table \@ref(tab:tinytableref)

```{r tinytableref}
library(tinytable)
tt(head(iris), caption = "(#tab:tinytableref) Hello world!") |> style_tt(color = "blue")
```
````


