
# Tiny Tables

```{=html}
<script id="MathJax-script" async src="https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"></script>
<script>
MathJax = {
  tex: {
    inlineMath: [['$', '$'], ['\\(', '\\)']]
  },
  svg: {
    fontCache: 'global'
  }
};
</script>
```

`tinytable` is a small but powerful `R` package to draw HTML, LaTeX, Word, PDF, Markdown, and Typst tables. The interface is minimalist, but it gives users direct and convenient access to powerful frameworks to create endlessly customizable tables.

Install the latest version from R-Universe or CRAN:

```{r, eval = FALSE}
install.packages("tinytable",
  repos = c("https://vincentarelbundock.r-universe.dev", "https://cran.r-project.org")
)
```

This tutorial introduces the main functions of the package. It is also [available as a single PDF document.](tinytable_tutorial.pdf)

Load the library and set some global options:

```{r}
library(tinytable)
options(tinytable_tt_digits = 3)
options(tinytable_latex_placement = "H")
```

Draw a first table:

```{r}
x <- mtcars[1:4, 1:5]
tt(x)
```

## Width and height

The `width` arguments indicating what proportion of the line width the table should cover. This argument accepts a number between 0 and 1 to control the whole table width, or a vector of numeric values between 0 and 1, representing each column.

```{r}
tt(x, width = 0.5)
```

```{r}
tt(x, width = 1)
```

We can control individual columns by supplying a vector. In that case, the sum of `width` elements determines the full table width. For example, this table takes 70% of available width, with the first column 3 times as large as the other ones.

```{r}
tt(x, width = c(.3, .1, .1, .1, .1))
```

When the sum of the `width` vector exceeds 1, it is automatically normalized to full-width. This is convenient when we only want to specify column width in relative terms:

```{r}
tt(x, width = c(3, 2, 1, 1, 1))
```

When specifying a table `width`, the text is automatically wrapped to appropriate size:


```{r}
lorem <- data.frame(
  Lorem = "Sed ut perspiciatis unde omnis iste natus error sit voluptatem accusantium doloremque laudantium, totam rem aperiam, eaque ipsa quae ab illo inventore veritatis et quasi architecto beatae vitae dicta sunt explicabo.",
  Ipsum = " Nemo enim ipsam voluptatem quia voluptas sit aspernatur aut odit aut fugit, sed quia consequuntur magni dolores eos."
)

tt(lorem, width = 3 / 4)
```

The `height` argument controls the height of each row in em units:

```{r}
tt(mtcars[1:4, 1:5], height = 3)
```


## Footnotes

The `notes` argument accepts single strings or named lists of strings:

```{r}
#| tbl-cap: "A full-width table with wrapped text in cells and a footnote."
n <- "Fusce id ipsum consequat ante pellentesque iaculis eu a ipsum. Mauris id ex in nulla consectetur aliquam. In nec tempus diam. Aliquam arcu nibh, dapibus id ex vestibulum, feugiat consequat erat. Morbi feugiat dapibus malesuada. Quisque vel ullamcorper felis. Aenean a sem at nisi tempor pretium sit amet quis lacus."

tt(lorem, notes = n, width = 1)
```

When `notes` is a named list, the names are used as identifiers and displayed as superscripts:

```{r}
tt(x, notes = list(a = "Blah.", b = "Blah blah."))
```

We can also add markers in individual cells by providing coordinates:

```{r}
tt(x, notes = list(
  a = list(i = 0:1, j = 1, text = "Blah."),
  b = "Blah blah."
))
```


## Captions and cross-references

In Quarto, one should always  specify captions cross-references using chunk options, and should *not* use the `caption` argument. This is because Quarto automatically post-processes tables, and may introduce conflict with the captions inserted by `tinytable`. For example:

````{verbatim}
@tbl-blah shows that...

```{r}
#| label: tbl-blah
#| tbl-cap: "Blah blah blah"
library(tinytable)
tt(mtcars[1:4, 1:4])
```
````

```{r, include=FALSE}
# knitr adds this by default, I guess, which duplicates [H][H]
options(tinytable_latex_placement = NULL)
```

And here is the rendered version of the code chunk above:

@tbl-blah shows that...

```{r}
#| label: tbl-blah
#| tbl-cap: "Blah blah blah"
library(tinytable)
tt(mtcars[1:4, 1:4])
```

One exception to the injunction above is when rendering a Quarto document to LaTeX using `theme_latex(multipage = TRUE, rowhead = 1)`. In that case, one must *avoid* using the Quarto chunk option, because these options trigger Quarto post-processing that will conflict with the `longtblr` environment used to split long tables across multiple pages.

The alternative is to use to refer to tables using standard LaTeX syntax: `\ref{tbl-ex-multipage}`. Then, use the `caption` argument in `tt()` to specify both the label and the caption:

```{r}
#| eval: false
tt(iris, caption = "Example table.\\label{tbl-ex-multipage}") |>
  theme_latex(multipage = TRUE, rowhead = 1)
```

For standalone tables in any format (i.e., outside Quarto), you can use the `caption` argument like so:

```{r, eval=FALSE}
tt(x, caption = "Blah blah.\\label{tbl-blah}")
```

```{r, include=FALSE}
# knitr adds this by default, I guess, which duplicates [H][H]
options(tinytable_latex_placement = "H")
```

## Output formats

`tinytable` can produce tables in HTML, Word, Markdown, LaTeX, Typst, PDF, or PNG format. An appropriate output format for printing is automatically selected based on (1) whether the function is called interactively, (2) is called within RStudio, and (3) the output format of the Rmarkdown or Quarto document, if applicable. Alternatively, users can specify the print format in `print()` or by setting a global option:

```{r}
#| eval: false
tt(x) |> print("markdown")
tt(x) |> print("html")
tt(x) |> print("latex")

options(tinytable_print_output = "markdown")
```

With the `save_tt()` function, users can also save tables directly to PNG (images), PDF or Word documents, and to any of the basic formats. All we need to do is supply a valid file name with the appropriate extension (ex: `.png`, `.html`, `.pdf`, etc.):

```r
tt(x) |> save_tt("path/to/file.png")
tt(x) |> save_tt("path/to/file.pdf")
tt(x) |> save_tt("path/to/file.docx")
tt(x) |> save_tt("path/to/file.html")
tt(x) |> save_tt("path/to/file.tex")
tt(x) |> save_tt("path/to/file.md")
```

`save_tt()` can also return a string with the table in it, for further processing in `R`. In the first case, the table is printed to console with `cat()`. In the second case, it returns as a single string as an `R` object.

```{r}
tt(mtcars[1:10, 1:5]) |>
  group_tt(
    i = list(
      "Hello" = 3,
      "World" = 8
    ),
    j = list(
      "Foo" = 2:3,
      "Bar" = 4:5
    )
  ) |>
  print("markdown")
```

```{r}
tt(mtcars[1:10, 1:5]) |>
  group_tt(
    i = list(
      "Hello" = 3,
      "World" = 8
    ),
    j = list(
      "Foo" = 2:3,
      "Bar" = 4:5
    )
  ) |>
  save_tt("markdown")
```

## Combination and exploration

Tables can be explored, modified, and combined using many of the usual base `R` functions:

```{r}
a <- tt(mtcars[1:2, 1:2])
a

dim(a)
ncol(a)
nrow(a)
names(a)
```

Tables can be combined with the usual `rbind()` function:

```{r}
a <- tt(mtcars[1:3, 1:2], caption = "Combine two tiny tables.")
b <- tt(mtcars[4:5, 8:10])

rbind(a, b)

rbind(a, b) |> format_tt(replace = "")
```

The `rbind2()` S4 method is slightly more flexible than `rbind()`, as it supports arguments `headers` and `use_names`.

Omit `y` header:

```{r}
rbind2(a, b, headers = FALSE)
```

Bind tables by position rather than column names:

```{r}
rbind2(a, b, use_names = FALSE)
```

## Select columns

The `subset()` function from base `R` can be used to select columns from a `tinytable`. This is especially useful when applying conditional styling based on column values, and then removing them. For example, if we have a table with 6 rows and three `Species`:

```{r}
dat <- do.call(rbind, by(iris, ~Species, head, n = 2))
dat
```

We highlight the `versicolor` rows in pink and remove the `Species` column:

```{r}
tt(dat) |>
  style_tt(Species == "versicolor", background = "pink") |>
  subset(select = -Species)
```

Or 

```{r}
tt(dat) |>
  style_tt(Species == "versicolor", background = "pink") |>
  subset(select = c(Sepal.Length, Sepal.Width))
```


## Rename columns

As noted above, `tinytable` tries to be standards-compliant, by defining methods for many base `R` functions. The benefit of this approach is that instead of having to learn a `tinytable`-specific syntax, users can rename columns using all the tools they already know:

```{r}
a <- tt(mtcars[1:2, 1:2])
names(a) <- c("a", "b")
a
```

In a pipe-based workflow, we can use the `setNames()` function from base `R`:

```{r} 
mtcars[1:2, 1:2] |>
  tt() |>
  setNames(c("a", "b"))
```

